Ant cuticle image classification using texture analysis: a comparative study

### 2022-06-18 paper/main.tex
draft

\author{%
    Noah Gardner\affil{1},
    John Paul Hellenbrand\affil{2},
    Anthony Phan\affil{1},
    Evan Zhu\affil{1},
    Zhiling Long\affil{4},
    Min Wang\affil{3},
    Clint Penick\affil{2},
    and Chih-Cheng Hung\affil{1}\corrauth
}%

@@

\begin{abstract}
    There is a large variety of ant species, and most species are diverse in
    terms of size, shape, behaviors, and especially cuticle textures. However,
    the significance of ant cuticle texture is not widely researched. Ant
    cuticle texture presumably provides some type of function, and therefore is
    useful to research for ecological applications and bioinspired designs. This
    research employs image texture analysis and deep machine learning to
    automatically group similar ant species based on morphological traits. We
    provide a comparative study of the performance of texture analysis methods
    on ant images. We evaluate the results of the classification methods with
    modern visualization techniques.
    \colorbox{green}{TODO: what is the result? experiments? data?}
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

\section{Introduction}
\colorbox{green}{TODO: refer to ant head images in a consistent way}

\colorbox{green}{TODO: refer to traditional/classical in a consistent way}

\colorbox{green}{TODO: change ideal and non-ideal to features identified by
    the biologist and to invalid features (background, etc.)}

@@

In image texture analysis methods, the general goal is to automatically
categorize an object into a set of objects with similar texture-based features.
Texture analysis has shown promising results in related fields, such as plant
identification \cite{boudra_plant_2018} \colorbox{green}{TODO: more references
    and examples}. With modern texture analysis methods, the categorization of ants
can be automated and the results can be used to study the influence of cuticle
texture on ant ecology.

Image texture analysis heavily depends on the spatial relationships among gray
levels of pixels \cite{hung_image_2019}. Hence, feature extraction method is
usually applied to the spatial relationship. The gray-level co-occurrence matrix
(GLCM) and local binary patterns (LBP) are two of methods which are based on the
spatial relationships to extracting features for this purpose
\cite{haralick_textural_1973, goos_gray_2000}. Traditional image texture
analysis can be divided into four categories: statistical, structural,
model-based, and transform-based methods \colorbox{green}{[references]}. In the
past decade, the deep learning network becomes the mainstream in image
classification and segmentation \cite{liu_bow_2019}. Compared with the
traditional method in which a kernel must be designed for extracting features,
the deep learning network can automatically extract the features by itself.

In this study, we used some of both approaches for our work on the ant images.

\colorbox{green}{\parbox{\dimexpr\textwidth-2\fboxsep}{TODO:Specifically, what
        approaches used? Why do we use image texture for ant images? What are
        the contributions?}}

\colorbox{green}{TODO: briefly discuss results from the research}

\section{Related Work}

\colorbox{green}{TODO: introduce the section}

\subsection{Sculpturing Identification}
Taxonomists have developed extensive terminology describing ant cuticle
sculpturing as it is often a useful diagnostic trait to distinguish between
closely related ant species \cite{blaimer_taxonomy_2019,fisher_ants_2007}. The
definitive text on ant cuticle terminology — \textit{The Glossary of Surface
    Sculpturing} — contains over 100 terms to describe the cuticle sculpturing
patterns of ants \cite{harris_glossary_1979}. There is often substantial overlap
among terms, and closely related cuticle patterns are likely to be functionally
similar. For example, the definition for “imbricate” is, “partly overlapping and
appearing like shingles on a roof or scales on a fish,” which is difficult to
distinguish from the definition of tesselate, “made up of squares like a chess
board, either in sculpturing or in color.”

@@

\colorbox{green}{TODO: how do we connect this section to image texture analysis?
    (why do we research this and why image texture) }

\subsection{Texture Analysis}
\colorbox{green}{TODO: A brief survey on image texture and cite Liu's
    IJCV2019 survey paper}

\subsection{Deep Learning for Texture Analysis}

In the past decade, deep learning using convolutional neural networks (CNNs)
\cite{krizhevsky_imagenet_2017} has emerged as the state-of-the-art technology
for image analysis. Following this trend, various CNN-based network
architectures have been designed to specifically characterize texture images.
Among these, the Fisher-vector CNN descriptor (FV-CNN) proposed by Cimpoi et al.
\cite{cimpoi_describing_2014} is widely accepted as one of the most important
pioneering works. It applies FV pooling to deep features obtained via a CNN
pre-trained using the ImageNet \cite{krizhevsky_imagenet_2017} to obtain encoded
features for texture classification. Although it is capable of achieving much
improved classification accuracy comparing to traditional hand-crafted texture
features, FV-CNN does not support an end-to-end learning, where feature
extraction, encoding, and classification are separated from each other. To
achieve an end-to-end learning, Zhang et al. \cite{zhang_visual_2018} proposed
the deep texture encoding network (DeepTEN), in which a novel texture encoding
layer is added to a standard CNN architecture. Then, Xue et al.
\cite{xue_deep_2018} constructed the deep encoding pooling network (DEP), which
improves over DeepTEN by integrating local spatial characteristics into the
texture representation. Based on DeepTEN and DEP, Hu et al. \cite{hu_multi_2019}
further developed the multi-level texture encoding and representation (MuLTER)
network, which embeds a learnable encoding module at each convolutional layer so
that encoding is performed for both low-level and high-level features, yielding
a multi-level texture representation.

Other network architectures for end-to-end texture learning are also available.
For example, in the deep multiple-attribute-perceived network (MAP-Net)
\cite{zhai_deep_2019}, multiple perceptual attributes are progressively learned
in a mutually reinforced manner through multiple branches. In the deep
structure-revealed network (DSR-Net)\cite{zhai_deep_2020}, inherent structural
representation for a texture pattern is obtained by employing a primitive
capturing module to learn spatial primitives and a dependency learning module to
capture the dependency among the primitives. In \cite{mao_deep_2021}, a residual
pooling layer consisting of a residual encoding module and an aggregation module
is used to generate discriminative features of low dimensions. In
\cite{peeples_histogram_2021}, a histogram layer is designed to compute local
spatial distribution of CNN features. In \cite{chen_deep_2021}, an innovative
aggregation module is presented to exploit statistical self-similarity across
layers. All these architectures customize the standard CNN structure to
accomplish the characterization of certain spatial, visual, or statistical
nature unique to texture images.

\subsection{Insect Classification}
In this section, we provide an overview of some insect classification methods.
Proposed insect classification methods seek to classify insects at different
hierarchical levels, such as species, genus, family, and order. Additionally,
some methods may classify insects at a combination of different hierarchical
levels. Insect classification methods can be applied to a variety of fields. In
agriculture, insect classification methods can be used to identify the presence
of pest insects in crops, which can inform crop managers in their choice of
pesticides and help prevent crop loss \cite{liu_pestnet_2019,
    kasinathan_machine_2021}.

@@

Lim et al. \cite{lim_performance_2017} apply a CNN-based algorithm for insect
classification. Lim et al. \cite{lim_performance_2017} classify a subset of
insect species and families based on the classes available in the ImageNet
dataset. ImageNet is a widely used dataset of images labeled by experts with
millions of images and thousands of categories \cite{deng_imagenet_2009}. In the
ImageNet dataset, there are some categories that specify the class of the insect
on a species level, \textit{e.g.} \textit{monarch butterfly} and \textit{ringlet
    butterfly} as well as some categories that specify the class of the insect on a
family level, \textit{e.g.} \textit{ant}, \textit{fly}, and \textit{bee}
\cite{imagenet_labels}. Lim et al. \cite{lim_performance_2017} use a modified
AlexNet architecture and experiment with different numbers of kernels and their
effect the performance of the model. Glick et al. \cite{glick_insect_2016}
employ a similar approach by classifying 277 insect classes from ImageNet using
a hierarchical convolutional neural network. The results from Lim et al.
\cite{lim_performance_2017} and Glick et al. \cite{glick_insect_2016} suggest
that a CNN is capable of differentiating between different hierarchical classes
of insects.

@@

\colorbox{green}{TODO: do not jump into subsection directly without any context}

@@

In this section, we describe the creation of the custom dataset used in this
research. In our dataset, we use ant head images from AntWeb
\cite{perrichot_antweb_2012} and define two categories for them based on the
appearance of the cuticle texture: \textit{rough} and \textit{smooth}. Some
randomly selected images from each category are shown in Figures
\ref{fig:rough-cuticle-texture} and \ref{fig:smooth-cuticle-texture}.

@@

% \subsubsection{Sculpture identification protocol}

@@

% \subsubsection{Data collection}

@@

% \subsubsection{Data preprocessing}

@@

\colorbox{green}{TODO: kviews, gabor, etc. will go here}

@@

\colorbox{green}{\parbox{\dimexpr\textwidth-2\fboxsep}{TODO: I suggest that
        at least one deep learning based texture classification algorithm be applied
        to the dataset for the comparative study. When this is the case, one more
        subsection should be included here to describe that/those algorithm(s).}}

@@

We evaluate the performance of the models according to standard evaluation
methods. Since we are working with a binary classification problem, we use a
standard confusion matrix to evaluate the accuracy, precision, and F1 score.
\colorbox{green}{(TODO: AUC and ROC can also be used as additional evaluation
    methods.)} We also apply Grad-CAM with manual inspection to visualize the
activation weights for classified images to visualize which features lead to the
classification result. Finally, we apply t-SNE to visualize the separation
learned for the model to further analyze the classifications made by the model.

@@

% vgg results
\subsection{VGG Models}

To begin, we share the results on the VGG model architectures on our custom
dataset. The classification results are summarized in a confusion matrix for
each model. Then, the statistics for each VGG architecture are shared in Table
\ref{tab:vgg_results}. The results are collected on 8 runs of training and
averaged. The results are rounded to 2 decimal places where appropriate.

\begin{table}[h]
    \centering
    \caption{Average results for VGG architectures on ant head image dataset.}
    \begin{tabular}{r|llll}
              & Recall & Precision & F1 Score & Accuracy \\
        \hline
        VGG11 & 0.83   & 0.88      & 0.85     & 0.80     \\
        VGG13 & 0.87   & 0.87      & 0.87     & 0.82     \\
        VGG16 & 0.86   & 0.86      & 0.86     & 0.80     \\
        VGG19 & 0.87   & 0.87      & 0.87     & 0.81     \\
    \end{tabular}
    \label{tab:vgg_results}
\end{table}

\subsection{ResNet Models}

Next, we share the results on the ResNet model architectures on our custom
dataset with random weight initialization. The classification results are
summarized in a confusion matrix for each model. Then, the statistics for each
ResNet architecture are shared in Table \ref{tab:resnet_results}. The results
are collected on 8 runs of training and averaged. The results are rounded
to 2 decimal places where appropriate.

\begin{table}[h]
    \centering
    \caption{Average results for ResNet architectures on ant head image dataset.}
    \begin{tabular}{r|llll}
                  & Recall & Precision & F1 Score & Accuracy \\
        \hline
        ResNet18  & 0.84   & 0.87      & 0.85     & 0.80     \\
        ResNet50  & 0.86   & 0.83      & 0.84     & 0.77     \\
        ResNet101 & 0.83   & 0.84      & 0.83     & 0.76     \\
    \end{tabular}
    \label{tab:resnet_results}
\end{table}

\clearpage
\subsection{Fine-tuned ResNet Models}

Next, we share the results on the fine-tuned ResNet model architectures on our
custom dataset with pretrained weights. The classification results are
summarized in a confusion matrix for each model. Then, the statistics for each
ResNet architecture are shared in Table \ref{tab:fresnet_results}. The results
are collected on 8 runs of training and averaged. The results are rounded
to 2 decimal places where appropriate.

\begin{table}[h]
    \centering
    \caption{Average results for Fine-tuned ResNet architectures on ant head image dataset.}
    \begin{tabular}{r|llll}
                  & Recall & Precision & F1 Score & Accuracy \\
        \hline
        ResNet18  & 0.89   & 0.94      & 0.91     & 0.88     \\
        ResNet50  & 0.89   & 0.92      & 0.91     & 0.87     \\
        ResNet101 & 0.89   & 0.94      & 0.92     & 0.88     \\
    \end{tabular}
    \label{tab:fresnet_results}
\end{table}

@@

The results in the previous sections show that the fine-tuned ResNet models
outperform the VGG and randomly initialized ResNet models on the task of ant
head image classification. It should be noted that due to the class imbalance in
the dataset, the F1 score is the preferable metric to the accuracy. On average,
the fine-tuned ResNet101 model performed the best with an average F1 score of
0.92. We further analyze the separation learned by both ResNet101 models in the
following section.

@@

\subsection{GradCAM Visualization}

In this section, we provide some visual analysis of some correctly and
incorrectly classified images using GradCAM. We provide two categories and two
subcategories in our analysis. The two categories are correct and incorrect
classification. Regardless of the feature activation map, the correctly
classified images have the same predicted label as the ground truth, and
incorrectly classified images have different predicted labels. The two
subcategories are ideal and non-ideal feature activation. In the ideal case, the
features that are used to compute the classification are the same as the
features used by the assistants in the sculpture identification process. In
general, the features used by the assistants are the textures of the cuticle on
the ant head. In the non-ideal case, the features used to compute the
classification are not from the head, for example, from the background,
extraneous text, or the body of the ant. We used randomly selected images from
the dataset and the fine-tuned ResNet101 model to perform the analysis. We show
the GradCAM results in Figures \ref{fig:correct_ideal},
\ref{fig:correct_nonideal}, \ref{fig:incorrect_ideal}, and
\ref{fig:incorrect_nonideal}. The left image shows the preprocessed image input
to the model. The right image shows the GradCAM output based on the
classification. Four specimens were selected randomly from each category and
subcategory.

Correctly classified images which use ideal features show the ideal performance
of the model. Incorrectly classified images which use ideal features should be
further analyzed. In essence, the model in this situation knows \textit{where}
to look, but not \textit{what} to look for. In Figure
\ref{fig:incorrect_ideal_84}, the features activated are mostly in the correct
location on the ant head, and the rough texture is clearly visible, yet the
model predicts the incorrect class \textit{smooth}. Similarly in Figure
\ref{fig:incorrect_ideal_177}, the features activated are also mostly in the
correct location, yet the model predicts the incorrect class \textit{smooth}. In
this case, it may be due to the pose of the ant being slightly different from
the average pose. In the incorrectly classified images with non-ideal features,
analysis shows that the model is unable to find \textit{where} to look, and
obtains feature information from other parts of the ant or the background. Cases
where the image was correctly classified using the non-ideal features can
basically be seen as noise. In order to further analyze this class, we should
introduce some parameter such as model confidence to examine further.

\newcommand{\subwidth}{0.35\textwidth}
\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/518.png}
        \caption{Ground truth: smooth}
        \label{fig:correct_ideal_518}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/808.png}
        \caption{Ground truth: smooth}
        \label{fig:correct_ideal_808}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/842.png}
        \caption{Ground truth: rough}
        \label{fig:correct_ideal_842}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/1091.png}
        \caption{Ground truth: rough}
        \label{fig:correct_ideal_1091}
    \end{subfigure}
    \caption{Correctly classified images using ideal features.}
    \label{fig:correct_ideal}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/346.png}
        \caption{Ground truth: rough}
        \label{fig:correct_nonideal_346}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1554.png}
        \caption{Ground truth: rough}
        \label{fig:correct_nonideal_1554}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1694.png}
        \caption{Ground truth: rough}
        \label{fig:correct_nonideal_1694}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/388.png}
        \caption{Ground truth: smooth}
        \label{fig:correct_nonideal_388}
    \end{subfigure}
    \caption{Correctly classified images using non-ideal features.}
    \label{fig:correct_nonideal}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/84.png}
        \caption{Ground truth: rough}
        \label{fig:incorrect_ideal_84}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/86.png}
        \caption{Ground truth: smooth}
        \label{fig:incorrect_ideal_86}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/177.png}
        \caption{Ground truth: smooth}
        \label{fig:incorrect_ideal_177}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/197.png}
        \caption{Ground truth: smooth}
        \label{fig:incorrect_ideal_197}
    \end{subfigure}
    \caption{Incorrectly classified images using ideal features.}
    \label{fig:incorrect_ideal}
\end{figure}

>>>>

\author{%
    Noah Gardner\affil{1},
    John Paul Hellenbrand\affil{2},
    Anthony Phan\affil{1},
    Haige Zhu\affil{1},
    Zhiling Long\affil{4},
    Min Wang\affil{3},
    Clint Penick\affil{2},
    and Chih-Cheng Hung\affil{1}\corrauth
}%

@@

\begin{abstract}
    There is a large variety of ant species, and most species are diverse in
    terms of size, shape, behaviors, and especially cuticle textures. However,
    the significance of ant cuticle texture is not widely researched. Ant
    cuticle texture presumably provides some type of function, and therefore is
    useful to research for ecological applications and bioinspired designs. This
    research employs image texture analysis and deep machine learning to
    automatically group similar ant species based on morphological traits. We
    provide a comparative study of the performance of general image
    classification and texture analysis methods on ant head images. We evaluate
    the results of the classification methods with modern visualization
    techniques. Our results show that deep learning methods are applicable for
    classifying ant cuticle textures. We also show that the performance of the
    deep texture analysis methods performs better than the deep classification
    methods.
\end{abstract}
\keywords{texture analysis; image processing; classification; machine learning;
    ant cuticle images; ecology}
\maketitle

\section{Introduction}

@@

Our primary contribution is the custom dataset used to explore the relationship
between ant cuticle texture and automated texture analysis methods. We compare
classical and modern image classification methods and texture analysis methods
by evaluating their performance on our dataset.

\section{Related work and literature review}

In this section we will review the literature for ant sculpturing identification
and texture analysis. Additionally, we discuss the work related to automated
insect identification.

\subsection{Sculpturing identification}
Taxonomists have developed extensive terminology describing ant cuticle
sculpturing as it is often a useful diagnostic trait to distinguish between
closely related ant species \cite{blaimer_taxonomy_2019,fisher_ants_2007}. The
definitive text on ant cuticle terminology — \textit{The Glossary of Surface
    Sculpturing} — contains over 100 terms to describe the cuticle sculpturing
patterns of ants \cite{harris_glossary_1979}.

@@

There is often substantial overlap among terms, and closely related cuticle
patterns are likely to be functionally similar. For example, the definition for
“imbricate” is, “partly overlapping and appearing like shingles on a roof or
scales on a fish,” which is difficult to distinguish from the definition of
tesselate, “made up of squares like a chess board, either in sculpturing or in
color.” However, we note that the terms used to describe the cuticle patterns
are similar to the terms used to describe textural patterns in texture analysis.
In the next subsection, we review some literature of image texture analysis.

\subsection{Texture analysis}

Image texture analysis is an important part of image interpretation as many
objects appearing in the textural patterns either partially or completely inside
an image. This phenomenon appears frequently in many images we encountered in
our environment as almost everybody is using the iPhone to capture the photos.
Scientists are also studying their subjects depending on taking the images
remotely such as remotely sensed satellite images. With an abundant number of
textural images surrounding us, it is essential to have a reliable automatic
tool for the textural image interpretation.

Classical computer methods are developed for textural images based on the
spatial relationship of gray levels of pixels encoded in the texture
\cite{haralick_image_1985,hung_image_2019}. Those methods by de-facto follow the
steps used in the statistical pattern recognition
\cite{fukunaga_introduction_2013}. Therefore, textural features are extracted,
and the classification and segmentation methods are used for the interpretation
either using the supervised or unsupervised approaches. The interpretation
method is usually taken and modified from the traditional pattern recognition
methods. For example, the gray-level co-occurrence matrix (GLCM) and local
binary patterns (LBP) are two of methods for extracting features
\cite{haralick_textural_1973, goos_gray_2000}.

Similar to the traditional pattern recognition, classical image texture analysis
methods can be grouped into four categories: statistical, structural,
model-based, and transform-based methods \cite{bharati_image_2004}. Among all of
the categories, the statistical method is often used by employing the method
from the pattern recognition principle.  In doing so, an image texture is
represented as feature vectors and then fed into the algorithm. Markov random
field models are also studied for textural image interpretation
\cite{hassner_use_1981, cross_markov_1983}. The transform-based methods use some functions to
decompose an image texture into a set of basic feature images. Gabor filters and
Wavelet expansions are two of the widely used approaches
\cite{bovik_multichannel_1990}.

While most feature vectors are established based on a single pixel, Liu et al.
gave a comprehensive survey on textural characterization in which they call this
feature extraction as texture representation \cite{liu_bow_2019}. Their survey
on Bag of Words and Convolutional Neural Networks (CNNs)
\cite{krizhevsky_imagenet_2017} are the spatial relationship concepts. The
K-views was developed for taking the spatial information in the clustering
approach \cite{hung_image_2019}.

In the past decade, deep learning using CNNs has emerged as the mainstream
technology for image analysis. Following this trend, various CNN-based network
architectures have been designed to specifically characterize textural images.
Among these, the Fisher-vector CNN descriptor (FV-CNN) proposed by Cimpoi et
al. \cite{cimpoi_deep_2015} is widely accepted as one of the most important
pioneering works. It applies FV pooling to deep features obtained via a CNN
pre-trained using the ImageNet \cite{krizhevsky_imagenet_2017} to obtain encoded features
for texture classification. Although it is capable of achieving much improved
classification accuracy comparing to traditional hand-crafted texture features,
FV-CNN does not support an end-to-end learning, where feature extraction,
encoding, and classification are separated from each other. To achieve an
end-to-end learning, Zhang et al. \cite{zhang_deep_2017} proposed the deep texture
encoding network (DeepTEN), in which a novel texture encoding layer is added to
a standard CNN architecture. Then, Xue et al. \cite{xue_deep_2018} constructed the
deep encoding pooling network (DEP), which improves over DeepTEN by integrating
local spatial characteristics into the texture representation. Based on DeepTEN
and DEP, Hu et al. \cite{hu_multi_2019} further developed the multi-level texture
encoding and representation (MuLTER) network, which embeds a learnable encoding
module at each convolutional layer so that encoding is performed for both
low-level and high-level features, yielding a multi-level texture
representation.

Other network architectures for end-to-end texture learning are also available.
For example, in the deep multiple-attribute-perceived network (MAP-Net)
\cite{zhai_deep_2019}, multiple perceptual attributes are progressively learned
in a mutually reinforced manner through multiple branches. In the deep
structure-revealed network (DSR-Net) \cite{zhai_deep_2020}, inherent structural
representation for a texture pattern is obtained by employing a primitive
capturing module to learn spatial primitives and a dependency learning module to
capture the dependency among the primitives. In \cite{mao_deep_2021}, a residual
pooling layer consisting of a residual encoding module and an aggregation module
is used to generate discriminative features of low dimensions. In
\cite{peeples_histogram_2021}, a histogram layer is designed to compute local
spatial distribution of CNN features. In \cite{chen_deep_2021}, an innovative
aggregation module is presented to exploit statistical self-similarity across
layers. All these architectures customize the standard CNN structure to
accomplish the characterization of certain spatial, visual, or statistical
nature unique to textural images.

Compared with the traditional method in which a kernel must be designed by an
engineer for extracting features, the deep learning networks can automatically
extract the features through the training. In addition, the deep learning
networks achieve the higher accuracy than that of the classical approaches,
although the deep networks require much more data for training.

\subsection{Insect classification}
In this section, we provide an overview of some insect classification methods.
Proposed insect classification methods seek to classify insects at different
hierarchical levels, such as species, genus, family, and order. Additionally,
some methods may classify insects at a combination of different hierarchical
levels. Insect classification methods can be applied to a variety of fields. In
agriculture, insect classification methods can be used to identify the presence
of pest insects in crops, which can inform crop managers in their choice of
pesticides and help prevent crop loss \cite{liu_pestnet_2019,
    kasinathan_machine_2021}.

@@

Lim et al. \cite{lim_performance_2017} apply a CNN-based algorithm for insect
classification. Lim et al. \cite{lim_performance_2017} classify a subset of
insect species and families based on the classes available in the ImageNet
dataset. ImageNet is a widely used dataset of images labeled by experts with
millions of images and thousands of categories \cite{deng_imagenet_2009}. In the
ImageNet dataset, there are some categories that specify the class of the insect
on a species level, \textit{e.g.} \textit{monarch butterfly} and \textit{ringlet
    butterfly} as well as some categories that specify the class of the insect on a
family level, \textit{e.g.} \textit{ant}, \textit{fly}, and \textit{bee}
\cite{deng_imagenet_2009}. Lim et al. \cite{lim_performance_2017} use a modified
AlexNet architecture and experiment with different numbers of kernels and their
effect the performance of the model. Glick et al. \cite{glick_insect_2016}
employ a similar approach by classifying 277 insect classes from ImageNet using
a hierarchical convolutional neural network. The results from Lim et al.
\cite{lim_performance_2017} and Glick et al. \cite{glick_insect_2016} suggest
that a CNN is capable of differentiating between different hierarchical classes
of insects.

@@

In this section, we provide an overview of the methodology used in this paper.
We describe the steps used to prepare the custom dataset. Finally, we describe
the algorithms applied and experimental setup to obtain the results.

@@

In this section, we describe the creation of the custom dataset used in this
research. In our dataset, we use ant head images from AntWeb
\cite{perrichot_antweb_2012} and define two categories for them based on the
appearance of the cuticle texture: \textit{rough} and \textit{smooth}. We also
refer to these ant head images, which describe the pose of the ant, as ant
cuticule images, due to the clear cuticle texture present on the head. Some
randomly selected images from each category are shown in Figures
\ref{fig:rough-cuticle-texture} and \ref{fig:smooth-cuticle-texture}.

@@

@@

@@

@@

The K-means algorithm is one of widely used clustering algorithms in the pattern
recognition \cite{lloyd_least_1982}. It is a single pixel based classification
algorithm for images. The K-views is an algorithm that uses several
characteristic views for the classification of images \cite{hung_use_2002}. The
K-views algorithm is suitable for classifying image textures that have basic
local patterns repeated in a periodic manner. In contrast to the K-means,
K-views looks at the neighboring pixels for providing the spatial relationship
for classification.  Several variations of the basic K-views model have been
proposed \cite{yang_image_2003, lan_improved_2010}.

For the feasibility study in this ant image datasets, we use both K-means and
K-views in our experiments as both are statistical clustering methods. The
experiments may indicate how well the statistical approach will do in this type
of textural images. In pre-processing the dataset, we apply the Gabor filter to
transform the gray-scale ant image into four Gabor-filtered channels, subtract
the gray-scale image from each channel, take the absolute value of the resulting
images, and then normalize them. This pre-processing will highlight some of
characteristics features such as edge-like in the image. This step is very
useful in feeding the datasets to the K-views algorithms. A 10-channel Gabor
filter is applied in the pre-processing step for the K-means algorithm.

@@

% \colorbox{green}{\parbox{\dimexpr\textwidth-2\fboxsep}{TODO: I suggest that at
%         least one deep learning based texture classification algorithm be
%         applied to the dataset for the comparative study. When this is the
%         case, one more subsection should be included here to describe
%         that/those algorithm(s).}}

@@

We evaluate the performance of the models according to standard evaluation
methods. Since we are working with a binary classification problem, we use a
standard confusion matrix to evaluate the accuracy, precision, and F1 score. We
also apply Grad-CAM with manual inspection to visualize the activation weights
for classified images to visualize which features lead to the classification
result. Finally, we apply t-SNE to visualize the separation learned for the
model to further analyze the classifications made by the model.

@@

\subsection{Evaluation results}

To begin, we share the results on the each algorithm specified in the
methodology section using the evaluation metrics described. The results are
shown in Table \ref{tab:results}. It should be noted that due to the class
imbalance in the dataset, the F1 score is the preferable metric to the accuracy.
Therefore, the results have been sorted by F1 score in descending order.

\begin{table}[h]
    \centering
    \caption{Results}
    \input{assets/results.tex}
    \label{tab:results}
\end{table}

@@

The results in the previous sections show that the fine-tuned ResNet models
outperform the VGG and randomly initialized ResNet models on the task of ant
head image classification. Additionally, the DRP models perform better than the
the general deep learning models VGG and ResNet on the task of ant head image
classification.The fine-tuned DRP single-layer model with auxiliary classifier
performed the best with an average F1 score of 0.92. We further analyze the
separation learned by both ResNet101 models in the following section.

% \colorbox{green}{(TODO: DRP is best performing model, it should be included in
%     visualization.)}

@@

\subsection{GradCAM visualization}

In this section, we provide some visual analysis of some correctly and
incorrectly classified images using GradCAM. We provide two categories and two
subcategories in our analysis. The two categories are correct and incorrect
classification. Regardless of the feature activation map, the correctly
classified images have the same predicted label as the ground truth, and
incorrectly classified images have different predicted labels. The two
subcategories are expected and non-expected feature activation. In the expected
case, the features that are used to compute the classification are the same as
the features used by the assistants in the sculpture identification process. In
general, the features used by the assistants are the textures of the cuticle on
the ant head. In the non-expected case, the features used to compute the
classification are not from the head, for example, from the background,
extraneous text, or the body of the ant. We used randomly selected images from
the dataset and the fine-tuned ResNet101 model to perform the analysis. We show
the GradCAM results in Figures \ref{fig:correct_ideal},
\ref{fig:correct_nonideal}, \ref{fig:incorrect_ideal}, and
\ref{fig:incorrect_nonideal}. The left image shows the preprocessed image input
to the model. The right image shows the GradCAM output based on the
classification. Four specimens were selected randomly from each category and
subcategory.

Correctly classified images which use the expected features show the ideal
performance of the model. Incorrectly classified images which use the expected
features should be further analyzed. In essence, the model in this situation
knows \textit{where} to look, but not \textit{what} to look for. In Figure
\ref{fig:incorrect_ideal_84}, the features activated are mostly in the correct
location on the ant head, and the rough texture is clearly visible, yet the
model predicts the incorrect class \textit{smooth}. Similarly in Figure
\ref{fig:incorrect_ideal_177}, the features activated are also mostly in the
correct location, yet the model predicts the incorrect class \textit{smooth}. In
this case, it may be due to the pose of the ant being slightly different from
the average pose. In the incorrectly classified images with non-expected
features, analysis shows that the model is unable to find \textit{where} to
look, and obtains feature information from other parts of the ant or the
background. Cases where the image was correctly classified using the
non-expected features can basically be seen as noise. In order to further
analyze this class, we should introduce some parameter such as model confidence
to examine further.

\newcommand{\subwidth}{0.35\textwidth}
\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/518.png}
        \caption{Ground truth: smooth}
        \label{fig:correct_ideal_518}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/808.png}
        \caption{Ground truth: smooth}
        \label{fig:correct_ideal_808}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/842.png}
        \caption{Ground truth: rough}
        \label{fig:correct_ideal_842}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_ideal/1091.png}
        \caption{Ground truth: rough}
        \label{fig:correct_ideal_1091}
    \end{subfigure}
    \caption{Correctly classified images using expected features.}
    \label{fig:correct_ideal}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/346.png}
        \caption{Ground truth: rough}
        \label{fig:correct_nonideal_346}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1554.png}
        \caption{Ground truth: rough}
        \label{fig:correct_nonideal_1554}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/1694.png}
        \caption{Ground truth: rough}
        \label{fig:correct_nonideal_1694}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/correct_nonideal/388.png}
        \caption{Ground truth: smooth}
        \label{fig:correct_nonideal_388}
    \end{subfigure}
    \caption{Correctly classified images using non-expected features.}
    \label{fig:correct_nonideal}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/84.png}
        \caption{Ground truth: rough}
        \label{fig:incorrect_ideal_84}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/86.png}
        \caption{Ground truth: smooth}
        \label{fig:incorrect_ideal_86}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/177.png}
        \caption{Ground truth: smooth}
        \label{fig:incorrect_ideal_177}
    \end{subfigure}
    \begin{subfigure}{\subwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/gradcam/incorrect_ideal/197.png}
        \caption{Ground truth: smooth}
        \label{fig:incorrect_ideal_197}
    \end{subfigure}
    \caption{Incorrectly classified images using expected features.}
    \label{fig:incorrect_ideal}
\end{figure}
